Cluster Difference Imaging Photometric Survey (CDIPS) II. TOI$\,$837: A Young Validated Saturn-Sized Planet in IC~2602

### 2020-07-21 paper/ms.tex
Update manuscript
Co-authored-by: Rafael Brahm <rbrahm@gmail.com>

\begin{abstract}
  We report the discovery of TOI 837.01 and its validation as a
  transiting planet.  We characterize the system using data from the
  NASA TESS mission, the ESA Gaia mission, ground-based photometry
  from El Sauce and ASTEP400, and spectroscopy from the CHIRON, FEROS,
  and Veloce spectrographs.  We find that TOI$\,$837 is a $T=9.9\,{\rm
  mag}$ G0/F9 dwarf in the southern open cluster IC 2602.  The star
  and planet are therefore $35^{+11}_{-5}$ million years old.
  Combining the transit photometry with a prior on the stellar
  parameters derived from the cluster color-magnitude diagram, we find
  that the planet orbits close to its star ($P = 8.3\,{\rm d}$) and is
  slightly smaller than Jupiter ($R_{\rm p} = 0.80^{+0.09}_{-0.03}
  \,R_{\rm Jup}$).  The transits are either grazing or nearly grazing.
  From radial velocity monitoring, we limit the minimum mass to less
  than 1.20 $M_{\rm Jup}$ (3$\sigma$).  Grazing transits are a cause
  for concern, as they are usually indicative of astrophysical false
  positive scenarios.  Our follow-up data show that such scenarios are
  highly unlikely.  Our combined multi-color photometry,
  high-resolution imaging, and radial velocities rule out hierarchical
  eclipsing binary scenarios.  Background eclipsing binary scenarios,
  though limited by speckle imaging, formally remain a 0.2\%
  possibility.  TOI 837.01 is therefore a validated adolescent
  exoplanet.  Further observations of its stellar obliquity,
  atmosphere, and mass should confirm its planetary nature beyond
  doubt, and could improve our understanding of how the physical and
  orbital properties of exoplanets change in time.
\end{abstract}

@@

The Gaia imaging, reduced to its point-source catalog, provides the
initial context for analyzing the TESS data.  Stars brighter than
$T=16$, as queried from the Gaia DR2 source catalog, are shown with
white circles in Figure~\ref{fig:scene}, overlaid on the TESS image.
Given its galactic latitude of $b=-6^\circ$, it is not surprising that
the field of \tn\ is crowded.  The resolved stars that were of
immediate concern for our false positive analysis were as follows.
\begin{itemize}
  %
  \item \tn\ $\equiv$ TIC 460205581 ($T=9.9$). The target star.
  %
  \item Star A $\equiv$ TIC 847769574 ($T=14.6$), $2.3''$ West. The
    proper motions and parallax of this star imply it is comoving with
    \tn\ and that the two stars are separated by $6.6\pm 0.1\,{\rm pc}$.
    Star A is therefore likely an \cn\ member, but unlikely to be a bound
    binary compnaion.
  %
  \item Star B $\equiv$ TIC 460205587 ($T=13.1$), $5.4''$ North.  The
    Gaia parallax implies this is a giant background star.
\end{itemize}
An additional source, TIC 847769581, is $4.9''$ from the target, but
too faint ($T=18.8$) to be the source of the observed transit signal.

@@

Reconaissance spectroscopic follow-up is an essential step in vetting
planet candidates.  Medium to high-resolution spectra enable physical
characterization of the star and therefore planet.  Reducing multiple
spectra to a radial velocity time-series can enable planet mass
measurements, and also set limits on the mass of any nearby
companions.  Finally, if there are close or bright companions,
reconaissance spectra can also reveal the presence of a secondary set
of stellar lines.

@@

To derive the stellar parameters, we matched the set of CHIRON spectra
against a library of observed spectra, previously obtained using the
Tillinghast Reflect Echelle Spectrograph
\citep[TRES,][]{furesz_tres_2008} on the 1.5\,m reflector at the Fred
Lawrence Whipple Observatory (FLWO), Arizona, USA, and classified
using the Stellar Parameter Classification pipeline
\citep{buchhave_hatp16b_class_2010}.  We found the best matching
stellar parameters to be $T_\mathrm{eff} = 5899 \pm 55$K, $\log g =
4.496 \pm 0.011$ dex, and $\mathrm{[Fe/H]} = -0.069 \pm 0.042$ dex.

@@

The above exercise ruled out the possibility that the observed
eclipses are caused by a stellar-mass object orbiting \tn.  The lack
of a linear trend in the radial velocities, particularly in the FEROS
data, further constrains the presence of a hierarchical binary system.
To derive limits for this second possibility, we placed a wide
log-uniform prior on the semi-amplitude and period ($\log K\ [{\rm
m\,s^{-1}}] \sim \mathcal{U}(1,10^5)$, $\log P\ [{\rm days}] \sim
\mathcal{U}(0.1, 10^{15})$).  We then fitted for the semi-amplitude,
period, time of conjunction, instrument offsets, and jitter
parameters.  We converted the resulting posterior in period and
semi-amplitude to minimum mass and semi-major axis assuming Kepler's
third law.  The resulting $3\sigma$ limits are shown in
Figure~\ref{fig:fpscenario}.
{\it LGB NOTE: the procedure in the preceding paragraph might be modified to
	instead convert the FEROS limit $|\dot{\gamma}| < 0.82\,{\rm m\,s}^{-1}\,{\rm
		day}^{-1}$ directly into a semi-amplitude limit.}

@@

The reason the above statement is an insufficient argument against
BEBs is that \tn\ is not a randomly selected star---it was selected
because it shows eclipses.  Given a foreground star that shows
eclipses, the probability of a background star being present is much
greater than for an arbitrary foreground star.  The relevant
populations need to be modelled at the Monte Carlo level.  We opt to
use \texttt{VESPA} to perform this population modelling
\citep{morton_efficient_2012,vespa_2015}.

\texttt{VESPA} calculates the false positive probability for a transit
signal as
\begin{equation}
  {\rm FPP} = 1 - P_{\rm pl},
\end{equation}
where in our case the probability that the signal comes from a planet,
$P_{\rm pl}$, is given by
\begin{align}
  P_{\rm pl} = 
  \frac{
    \mathcal{L}_{\rm pl}\pi_{\rm pl}
  }{
    \mathcal{L}_{\rm pl}\pi_{\rm pl} + \mathcal{L}_{\rm BEB}\pi_{\rm BEB}
  },
\end{align}
where $\mathcal{L}_i$ is the model likelihood for the planet and BEB
scenarios, and $\pi_i$ is the model prior.  The terms labelled as
``BEB'' usually include other false positive scenarios (HEBs and EBs),
but our followup data have excluded these possibilities.  The priors
are evaluated using a combination of galactic population synthesis
\citep{girardi_star_2005}, binary star statistics
\citep{raghavan_survey_2010}, and specific planet occurrence rates
\citep[][Section~3.4]{morton_efficient_2012}.  The likelihoods are
evaluated by forward-modelling a representative population of
eclipsing bodies for model class, in which each population member has
a particular trapezoidal eclipse depth, total duration, and ingress
duration.  The likelihood is then calculated by multiplying the
probability distribution function of simulated population's shape
parameters with the posterior probability of the actual observed
eclipse shape.

@@

\section{System Modelling}
\label{sec:system}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.45\textwidth]{f9a.pdf}
		}

		\vspace{-0.5cm}
		\subfloat{
			\includegraphics[width=0.45\textwidth]{f9b.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Youth diagnostics.}
    {\it Top:} Lithium 6708\AA\ equivalent widths for \tn, field
    stars, and young open clusters.  The field star sample is drawn
    from Kepler planet hosts, and was measured by
    \citet{berger_identifying_2018} using Keck-HIRES.  The young open
    clusters members were surveyed by \citet{randich_gaiaeso_2018}
    using the UVES and GIRAFFE spectrographs at the ESO VLT.
    \citet{randich_gaiaeso_2018} found lithium depletion boundary ages
    for these clusters of $37.7^{+5.7}_{-4.8}\,{\rm Myr}$
    (NGC$\,$2547) and $43.7^{+4.3}_{-3.9}\,{\rm Myr}$ (\cn).  {\it
    Bottom:} Rotation periods for \tn\ and selected open clusters.
    The Pleiades (120$\,$Myr), Praesepe (670$\,$Myr), and NGC$\,$6811
    (1000$\,$Myr) are shown.  Their rotation periods were measured by
    \citet{rebull_rotation_2016a,douglas_poking_2017,douglas_k2_2019},
    and \citet{curtis_temporary_2019}, respectively.
    \label{fig:lithium_rotation}
	}
\end{figure}

@@

Figure~\ref{fig:hr} shows a Hertzsprung-Russell diagram of \tn, the
\cn\ cluster, and the ``neighborhood'' of spatially nearby stars.
Stars labelled as ``cluster members'' are those reported by
\citet{cantatgaudin_gaia_2018} based on Gaia DR2 positions, proper
motions, and parallaxes.  We included candidate members with formal
membership probability exceeding 10\%.  Most members appear to be
young and coeval.  \tn\ is in its expected position relative to the
other members along the cluster isochrone.  This photometrically
limits the presence of binary companions in the \tn\ system, to within
perhaps half the brightness ($\approx0.75$ magnitudes) of the target
star.

@@

We also fitted the available time-series photometry of \pn\ using two
different approaches.  To derive the most precise possible ephemeris,
we fitted the ground and space-based transits simultaneously.  To the
derive the physical parameters of the planet, we fitted the TESS data
alone, due to our better understanding of the underlying systematic
trends and the higher precision.

@@

A.J.\ and R.B.\ acknowledge support from project IC120009 ``Millennium
Institute of Astrophysics (MAS)'' of the Millenium Science Initiative,
Chilean Ministry of Economy. A.J.\ acknowledges additional support
from FONDECYT project 1171208.  J.I.V\ acknowledges support from
CONICYT-PFCHA/Doctorado Nacional-21191829.
%

C.Z.\ is supported by a Dunlap Fellowship at the Dunlap Institute for
Astronomy \& Astrophysics, funded through an endowment established by
the Dunlap family and the University of Toronto.
%

>>>>

\begin{abstract}
  We report the discovery of TOI 837.01 and its validation as a
  transiting planet.  We characterize the system using data from the
  NASA TESS mission, the ESA Gaia mission, ground-based photometry
  from El Sauce and ASTEP400, and spectroscopy from the CHIRON, FEROS,
  and Veloce spectrographs.  We find that TOI$\,$837 is a $T=9.9$ mag
  G0/F9 dwarf in the southern open cluster IC 2602.  The star
  and planet are therefore $35^{+11}_{-5}$ million years old.
  Combining the transit photometry with a prior on the stellar
  parameters derived from the cluster color-magnitude diagram, we find
  that the planet has an orbital period of $8.3\,{\rm d}$ and is
  slightly smaller than Jupiter ($R_{\rm p} = 0.80^{+0.09}_{-0.03}
  \,R_{\rm Jup}$).  The transits are either grazing or nearly grazing.
  From radial velocity monitoring, we limit the minimum mass to less
  than 1.20 $M_{\rm Jup}$ (3$\sigma$).  Grazing transits are a cause
  for concern, as they are usually indicative of astrophysical false
  positive scenarios.  Our follow-up data show that such scenarios are
  highly unlikely.  Our combined multi-color photometry,
  high-resolution imaging, and radial velocities rule out hierarchical
  eclipsing binary scenarios.  Background eclipsing binary scenarios,
  though limited by speckle imaging, formally remain a 0.2\%
  possibility.  TOI 837.01 is therefore a validated adolescent
  exoplanet.  Further observations of its stellar obliquity,
  atmosphere, and mass should confirm its planetary nature beyond
  doubt, and could improve our understanding of how the physical and
  orbital properties of exoplanets change in time.
\end{abstract}

@@

The Gaia imaging, reduced to its point-source catalog, provides the
initial context for analyzing the TESS data.  Stars brighter than
$T=16$, as queried from the Gaia DR2 source catalog, are shown with
white circles in Figure~\ref{fig:scene}, overlaid on the TESS image.
Given its galactic latitude of $b=-6^\circ$, it is not surprising that
the field of \tn\ is crowded.  The resolved stars that were of
immediate concern for our false positive analysis were as follows.
\begin{itemize}
  %
  \item \tn\ $\equiv$ TIC 460205581 ($T=9.9$). The target star.
  %
  \item Star A $\equiv$ TIC 847769574 ($T=14.6$), $2.3''$ West. The
    proper motions and parallax of this star imply it is co-moving with
    \tn\ and that the two stars are separated by $6.6\pm 0.1\,{\rm pc}$.
    Star A is therefore likely an \cn\ member, but unlikely to be a bound
    binary companion.
  %
  \item Star B $\equiv$ TIC 460205587 ($T=13.1$), $5.4''$ North.  The
    Gaia parallax implies this is a giant background star.
\end{itemize}
An additional source, TIC 847769581, is $4.9''$ from the target, but
too faint ($T=18.8$) to be the source of the observed transit signal.

@@

Reconnaissance spectroscopic follow-up is an essential step in vetting
planet candidates.  Medium to high-resolution spectra enable physical
characterization of the star and therefore planet.  Reducing multiple
spectra to a radial velocity time-series can enable planet mass
measurements, and also set limits on the mass of any nearby
companions.  Finally, if there are close or bright companions,
reconnaissance spectra can also reveal the presence of a secondary set
of stellar lines.

@@

To derive the stellar parameters, we matched the set of CHIRON spectra
against a library of observed spectra, previously obtained using the
Tillinghast Reflect Echelle Spectrograph
\citep[TRES,][]{furesz_tres_2008} on the 1.5\,m reflector at the Fred
Lawrence Whipple Observatory (FLWO), Arizona, USA, and classified
using the Stellar Parameter Classification pipeline
\citep{buchhave_hatp16b_class_2010}.  We found the best matching
stellar parameters to be $T_\mathrm{eff} = 5899 \pm 55$K, $\log g =
4.496 \pm 0.011$ dex, and $\mathrm{[Fe/H]} = -0.069 \pm 0.042$ dex.
We ultimately adopted a different set of stellar parameters for 
our analysis (see Section~\ref{subsec:starparams}).

@@

The above exercise ruled out the possibility that the observed
eclipses are caused by a stellar-mass object orbiting \tn.  The lack
of a linear trend in the radial velocities, particularly in the FEROS
data, further constrains the presence of a hierarchical binary system.
The consistency of the Gaia DR2 RV ($17.44 \pm 0.64\,$\kms) with
the FEROS RVs ($18.0 \pm 0.1$\kms) in theory places an even stronger limit on 
linear trends, since the two observation epochs are separated by roughly
five years.
To derive limits for this second possibility, we placed a wide
log-uniform prior on the semi-amplitude and period ($\log K\ [{\rm
m\,s^{-1}}] \sim \mathcal{U}(1,10^5)$, $\log P\ [{\rm days}] \sim
\mathcal{U}(0.1, 10^{15})$).  We then fitted for the semi-amplitude,
period, time of conjunction, instrument offsets, and jitter
parameters.  We converted the resulting posterior in period and
semi-amplitude to minimum mass and semi-major axis assuming Kepler's
third law.  The resulting $3\sigma$ limits are shown in
Figure~\ref{fig:fpscenario}.
{\it LGB NOTE: the procedure in the preceding paragraph might be modified to
	instead convert the FEROS limit $|\dot{\gamma}| < 0.82\,{\rm m\,s}^{-1}\,{\rm
		day}^{-1}$ directly into a semi-amplitude limit.}

@@

The reason the above statement is an insufficient argument against
BEBs is that \tn\ is not a randomly selected star---it was selected
because it shows eclipses.  Given a foreground star that shows
eclipses, the probability of a background star being present is much
greater than for an arbitrary foreground star.  The relevant
populations need to be modeled at the Monte Carlo level.  We opt to
use \texttt{VESPA} to perform this population modeling
\citep{morton_efficient_2012,vespa_2015}.

\texttt{VESPA} calculates the false positive probability for a transit
signal as
\begin{equation}
  {\rm FPP} = 1 - P_{\rm pl},
\end{equation}
where in our case the probability that the signal comes from a planet,
$P_{\rm pl}$, is given by
\begin{align}
  P_{\rm pl} = 
  \frac{
    \mathcal{L}_{\rm pl}\pi_{\rm pl}
  }{
    \mathcal{L}_{\rm pl}\pi_{\rm pl} + \mathcal{L}_{\rm BEB}\pi_{\rm BEB}
  },
\end{align}
where $\mathcal{L}_i$ is the model likelihood for the planet and BEB
scenarios, and $\pi_i$ is the model prior.  The terms labeled as
``BEB'' usually include other false positive scenarios (HEBs and EBs),
but our followup data have excluded these possibilities.  The priors
are evaluated using a combination of galactic population synthesis
\citep{girardi_star_2005}, binary star statistics
\citep{raghavan_survey_2010}, and specific planet occurrence rates
\citep[][Section~3.4]{morton_efficient_2012}.  The likelihoods are
evaluated by forward-modeling a representative population of
eclipsing bodies for model class, in which each population member has
a particular trapezoidal eclipse depth, total duration, and ingress
duration.  The likelihood is then calculated by multiplying the
probability distribution function of simulated population's shape
parameters with the posterior probability of the actual observed
eclipse shape.

@@

\section{System Modeling}
\label{sec:system}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=0.45\textwidth]{f9b.pdf}
		}

		\vspace{-0.5cm}
		\subfloat{
			\includegraphics[width=0.45\textwidth]{f9a.pdf}
		}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Youth diagnostics.}
	{\it Top:} Rotation periods for \tn\ and selected open clusters.
    The Pleiades (120$\,$Myr), Praesepe (670$\,$Myr), and NGC$\,$6811
    (1000$\,$Myr) are shown.  Their rotation periods were measured by
    \citet{rebull_rotation_2016a,douglas_poking_2017,douglas_k2_2019},
    and \citet{curtis_temporary_2019}, respectively.
    {\it Bottom:} Lithium 6708\AA\ equivalent widths for \tn, field
    stars, and young open clusters.  The field star sample is drawn
    from Kepler planet hosts, and was measured by
    \citet{berger_identifying_2018} using Keck-HIRES.  The young open
    clusters members were surveyed by \citet{randich_gaiaeso_2018}
    using the UVES and GIRAFFE spectrographs at the ESO VLT.
    \citet{randich_gaiaeso_2018} found lithium depletion boundary ages
    for these clusters of $37.7^{+5.7}_{-4.8}\,{\rm Myr}$
    (NGC$\,$2547) and $43.7^{+4.3}_{-3.9}\,{\rm Myr}$ (\cn).  
    \label{fig:lithium_rotation}
	}
\end{figure}

@@

Figure~\ref{fig:hr} shows a Hertzsprung-Russell diagram of \tn, the
\cn\ cluster, and the ``neighborhood'' of spatially nearby stars.
Stars labeled as ``cluster members'' are those reported by
\citet{cantatgaudin_gaia_2018} based on Gaia DR2 positions, proper
motions, and parallaxes.  We included candidate members with formal
membership probability exceeding 10\%.  Most members appear to be
young and coeval.  \tn\ is in its expected position relative to the
other members along the cluster isochrone.  This photometrically
limits the presence of binary companions in the \tn\ system, to within
perhaps half the brightness ($\approx0.75$ magnitudes) of the target
star.

@@

We also fitted the available time-series photometry of \pn\ using two
different approaches.  To derive the most precise possible ephemeris,
we fitted the ground and space-based transits simultaneously.  To
derive the physical parameters of the planet, we fitted the TESS data
alone, due to our better understanding of the underlying systematic
trends and the higher precision.

@@

A.J.\ and R.B.\ acknowledge support from project IC120009 ``Millennium
Institute of Astrophysics (MAS)'' of the Millenium Science Initiative,
Chilean Ministry of Economy. A.J.\ acknowledges additional support
from FONDECYT project 1171208.  J.I.V\ acknowledges support from
CONICYT-PFCHA/Doctorado Nacional-21191829.
R.B.\ acknowledges support from FONDECYT Post-doctoral Fellowship Project 3180246.
%
C.Z.\ is supported by a Dunlap Fellowship at the Dunlap Institute for
Astronomy \& Astrophysics, funded through an endowment established by
the Dunlap family and the University of Toronto.
%
